\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization

### 2010-09-28 honors_thesis/gecco_paper/paper.tex

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \includegraphics[width=.3\columnwidth]{messages1.eps}
  \caption{The messages needed to produce $\p_{t}$.  Items without an arrow
  pointing to them are messages the particle needs to receive.}
  \label{fig:messages1}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \includegraphics[width=.5\columnwidth]{messages2.eps}
  \caption{The messages needed to produce $\nonbest{\p}_{t+1}$.  Items without
  an arrow pointing to them are messages the particle needs to receive.}
  \label{fig:messages2}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \includegraphics[width=.6\columnwidth]{messages3.eps}
  \caption{The messages needed to produce $\nonbest{\nset}_{t+1}$.  Items
  without an arrow pointing to them are messages the particle needs to receive.
  $\nnset$ is the set of neighbors for each particle $\n$, and $\nsset$ is the
  set of $\n$'s speculative children.  Note the similarity between this and
  \fig{messages2}.}
  \label{fig:messages3}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \includegraphics[width=.8\columnwidth]{messages4.eps}
  \caption{The messages needed to produce $\p_{t+1}$.  Items
  without an arrow pointing to them are messages the particle needs to receive.
  Note that this is just a combination of \fig{messages2} and \fig{messages3}.}
  \label{fig:messages4}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \psfrag{nt1}[C][C]{$\nset_{t+1}$}
  \psfrag{nt2-e}[C][C]{$\noeval{\nset}_{t+2}$}
  \psfrag{pt2-e}[C][C]{$\noeval{\p}_{t+2}$}
  \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
  \includegraphics[width=.5\columnwidth]{messages5.eps}
  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
  and $\nset_{t+1}$ are both produced as in \fig{messages4}.}
  \label{fig:messages5}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere1.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:sphere1}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed themselves.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:griewank1}
\end{figure}

\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
runs found the global optimum, 0, while some did not.  A more accurate plot for
this function is one showing what percentage of the runs have found the global
optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
30 performing the PSO algorithm gets stuck a little less than half of the time,
while 240 particles provides enough exploration to find the global optimum
every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
particles greatly improves the chance of success with Griewank, while also
finding the global optimum in far fewer rounds of evaluations than standard
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

>>>>

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \includegraphics[width=.3\columnwidth]{messages1}
  \caption{The production of $\p_{t}$ from the original particle
  $\nonbest{\p}_{t}$ and the messages $\nonbest{\nset}_{t}$.}
  \label{fig:messages1}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \includegraphics[width=.5\columnwidth]{messages2}
  \caption{The production of $\nonbest{\p}_{t+1}$ from the original particle 
  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$ and
  $\nonbest{\sset}_{t+1}$, and intermediate particles.}
  \label{fig:messages2}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \includegraphics[width=.6\columnwidth]{messages3}
  \caption{The production of each $\nonbest{\n}_{t+1}$ from the original
  particle $\nonbest{\n}_{t}$, messages $\nonbest{\nnset}_{t}$ and
  $\nonbest{\nsset}_{t+1}$, and intermediate particles.  $\nnset$ is the set of
  neighbors for each particle $\n$, and $\nsset$ is the set of $\n$'s
  speculative children.  Note the similarity between this and \fig{messages2}.}
  \label{fig:messages3}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \includegraphics[width=.8\columnwidth]{messages4}
  \caption{The production of $\p_{t+1}$ from the original particle 
  $\nonbest{\p}_{t}$, messages $\nonbest{\nset}_{t}$, $\nonbest{\sset}_{t+1}$,
  $\nonbest{\nnset}_{t}$, and $\nonbest{\nsset}_{t+1}$, and intermediate
  particles.  Note that this is just a combination of \fig{messages2} and
  \fig{messages3}.}
  \label{fig:messages4}
\end{figure}

@@

\begin{figure}
  \centering
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \psfrag{nt1}[C][C]{$\nset_{t+1}$}
  \psfrag{nt2-e}[C][C]{$\noeval{\nset}_{t+2}$}
  \psfrag{pt2-e}[C][C]{$\noeval{\p}_{t+2}$}
  \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
  \includegraphics[width=.5\columnwidth]{messages5}
  \caption{The production of $\noeval{\p}_{t+2}$ and $\noeval{\sset}_{t+3}$
  from $\p_{t+1}$ and $\n_{t+1}$, each of which are produced as in
  \fig{messages4}.}
  \label{fig:messages5}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere1}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:sphere1}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optimum.  Griewank has a global optimum with a value of 0, and
sometimes the swarm finds the optimum and sometimes it doesn't.  Instead of
showing average function value at each iteration, a more accurate plot for
Griewank shows the percent of runs that have found the global optimum by each
iteration.  We show results in \fig{griewank1} for swarms of size 30 and 240
using the Ring topology.  One can see in the figure that both SEPSO algorithms
find the optimum faster than the original PSO.  However, because the swarm size
is so small, SEPSO gets stuck a little less than half of the time.
Surpisingly, SEPSO Pick Best improves the chance of success with Griewank,
while also finding the optimum faster.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, showing percent successful by each iteration instead of average
  function value.}
  \label{fig:griewank1}
\end{figure}

Because our speculative algorithm got stuck with such a small swarm size, we
ran a set of experiments on Griewank with 800 processors, giving swarms of size
100 and 800 instead of 30 and 240.  \fig{griewank2} shows the results.
\fig{griewank2} is very similar to \fig{griewank1}, except the speculative
algorithms find the global optimum essentially 100\% of the time.  Thus with a
few more processors, our algorithms find the optimum two to three times faster
than the original PSO without a significant loss of accuracy.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}
